Using Graphical Models for Novelty Detection on Semantic Room Classification

### 2011-05-09 semantic-nd/article.tex
Added stuff specially on results section.

In the presented case the conditional probability is approximated by the graphical model $G$
produced by the conceptual layer.
The \autoref{fig:conditional-prob-graph} represents a possible graph $G$ built from the conceptual
layer to represent the conditional probability on the sensed variables $x$.
$M_G(x)$ is used as an approximation for $P(x|\overline{novel})$.

@@

For interesting and practical reasons it was assumed that the unconditional distribution
could be modelled with all the variables independent from each other.
Which translates as $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate the probability of each feature $x_i$.
\autoref{fig:semisupervised-threshold} illustrates a graph $I$ to model independent variables.

@@

In order to verify the improvements obtained by using unlabelled data, a fictitious simple dataset was generated.
The fictitious distribution assumes that an independent and variable size set of features $x$ is generated by a
given room category category. In whole there was 11 different room categories and 7 different measured properties types.

The known data was extracted by generating samples from 6 of the 11 classes. Together with the generated known data, the
room category that generated is given as label. Using that we built the factors for the generation of graphs representing
the conditional distribution $G$.
Using unlabelled data we trained also a graph $I$ representing an unconditional distribution where all features are independent.

@@

Since the distribution is fictitious we were able to create an optimal novelty detector, as we can exactly measure $P(novel|x)$.
That served as a comparison on how optimal a detector could be.
A ROC curve was plotted for the three methods on \autoref{fig:fictitious-roc}.
As possible to see a correct estimation on the unconditional probability is important for obtaining a correct ordering of the
input space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\begin{itemize}
\item Show a correct estimation of unconditional probability to be important in estabilishing a novelty detection threshold
\item Show unlabelled data can be used for it.
\item Although focus on Room Categorization we expect the results and techniques to be usable on other areas.
\end{itemize}

On this paper it was presented how to define a stable novelty threshold function on top of
\emph{probabilistic graphical models} instantiated dynamically from sensed semantic data and
conceptual knowledge.

It was also shown that unlabelled data can be used to approximate unconditional probability and
improve a novelty threshold.

Quotting Andrzej: """ Being able to automatically explain why a class is novel and what
makes it different from any of the other classes -- This is super important for communication with the user and active learning of new room categories."""

>>>>

In the presented case the conditional probability is approximated by the graphical model $G$
produced by the conceptual layer.
The \autoref{fig:conditional-prob-graph} represents a possible graph $G$ built from the conceptual
layer to represent the conditional probability on the sensed variables $x$.
$P_G(x)$ is used as an approximation for $P(x|\overline{novel})$.

@@

For interesting and practical reasons it was assumed that the unconditional distribution
could be modelled with all the variables independent from each other.
Which translates as $P(x)=\prod{P_{x_i}(x_i)}$.
From the unlabelled data only the probability of each feature $x_i$ needs to be estimated.
\autoref{fig:semisupervised-threshold} illustrates a graph $I$ to model independent variables.

@@

In order to verify the improvements obtained by using unlabelled data, a fictitious simple
dataset was generated.
The fictitious distribution assumes that an independent and variable size set of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 7 different measured properties types.

The known data was extracted by generating samples from 6 of the 11 classes.
Together with the generated known data, the room category that generated is given as label.
That known data is used to train factors to represent a conditional distribution.
Using unlabelled data, factors are trained to approximate an unconditional distribution
where all features are independent.
\autoref{fig:simple-experiment} shows the graph structure used for approximate the conditional and
unconditional distributions trained from the labelled and unlabelled data.

\subfloat[Graph structure used to model conditional probability graph.
          The graph factors are trained using the labelled data.]{... conditional probability}
\qquad
\subfloat[Graph structure used to model unconditional probability graph.
          The graph factors are trained using the unlabelled data.]{... unconditional probability}

\caption{\label{fig:simple-experiment}The graph structures $G$ and $I$ used to model the
         conditional and unconditional probability.
         Note that graph $G$ correctly models the structure of the fictitious distribution, though
         during training it has not seen all possible room categories represented by variable $R_1$.}

\end{figure}

@@

Since the distribution is fictitious its possible to create an optimal novelty detector, as
its the error rate $P(novel|x)$ can be exactly measured.
That serves as a comparison on how optimal a detector can be using the same input space.
A ROC curve was plotted for the three methods on \autoref{fig:fictitious-roc}.
As possible to see the unconditional probability plays an important role in obtaining a correct
ordering of the input space.

The importance of a correct estimate of unconditional probability is expected to decrease as the
input space increases and allows the several existing classes to become easily distinguished.

For comparison ROC curves where plot for the developed threshold based on the number of input features.

\begin{figure}
\centering

\subfloat[3 sensed variables]{5 sensed variables}
\qquad
\subfloat[5 sensed variables]{5 sensed variables}

\subfloat[10 sensed variables]{10 sensed variables}
\qquad
\subfloat[50 sensed variables]{50 sensed variables}

\caption{\label{fig:fictitious-roc-breakdown}ROC curves plotted showing performance of the
         presented novelty detection method on graphs generated for different amount of
         sensed variables.
         As possible to see as the system gains more semantic information it becomes easier
         to detect novelty and the effect of picking an approximation for the unconditional
         probability decreases.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\begin{itemize}
\item Show a correct estimation of unconditional probability to be important in establishing a novelty detection threshold
\item Show unlabelled data can be used for it.
\item Although focus on Room Categorization we expect the results and techniques to be usable on other areas.
\end{itemize}

On this paper it was presented how to define a stable novelty threshold function on top of
\emph{probabilistic graphical models} instantiated dynamically from sensed semantic data and
conceptual knowledge.
The presented technique is based on the ratio between a conditional and unconditional probability.

It was shown that unlabelled data can be used to approximate unconditional probability and
improve a novelty threshold.
And that a correct estimation of unconditional probability plays an important role specially on
small input spaces.

Future will work will focus on how to explain why a class is novel and what makes it different
other classes.
That will lead to generation of useful information that can be used on communication with the user
and perform active learning of new room categories.
